Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-20 paper.tex
Fix some issues with polytomies

% Our second example shows a different aspect of this robustness to missing 
% data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
% obtained when we compute the likelihood of an example simulation,
% in which the ARG is at three different levels of completeness
% (illustrated in Fig~\ref{fig:lik-surface}B).
% We first run a simulation under the SMC
% using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
% ARG explicitly identifies all of the recombination and common 
% ancestor events that occurred in the simulation, and 
% we can therefore compute a likelihood under the HKY formulation
% (for example, using the \texttt{msprime.log\_arg\_likelihood} 
% function). The example full ARG in 
% Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
% unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
% left and right parents of the recombination event causing the 
% breakpoint at position 536~\cite[we need two nodes to uniquely identify
%  the passage of ancestral material through recombination 
% events; see][]{baumdicker_efficient_2021}, and therefore present
% in only those trees. 
% Node 8 is a coalescent node in the right-hand tree, but unary
% in the left-hand tree, marking the passage of the lineage through
% this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
% Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
% remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
% and compute the likelihood using eq.~\eqref{eq:full-lik},
% there is very little difference in the likelihood ratio
% curve from the exact value with complete input. 
% Most importantly, the minimum occurs at the true value of the 
% recombination parameter.
% When we further simplify
% the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
% we obtain an ARG in the standard format used by \texttt{msprime}.
% Although there is substantial information about the correlation
% structure of trees in this representation, we lose important
% information about recombination. This is demonstrated in 
% Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
% is clearly biased.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under the Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices:
    from 0 to 1,000 generations ago $N_e$ is 10,000;
    from 1,000 to 5,000 generations ago it is 5,000;
    and past that it is 10,000.
    Remaining parameters: genome length is 10kb, 20 diploid samples, and
    $r=2.5\times10^{-8}$. 
    There are 148 edges in the resulting ARG.
    A single $N_e$ value was estimated for each
    of the three slices of the ARG separated by
    the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. The number of edges per time slice
    (from most recent to oldest) was: 68, 57, and 23. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

% slicing of the ARG
Our final example explores decomposition of the likelihood
into separate components for different segments of time.
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, with a known time to recombination,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the left or
right hand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.
% This approach therefore strengthens
% the inherent ability of ARGs to provide us with temporal
% resolution and context to a dataset of genetic variation.
% It also opens up the opportunity for more efficient optimization
% strategies during inference given that each slice of the ARG can be
% processed independently.

>>>>

Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably
robust to the presents of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and 
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because 
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent 
is not unary. Thus, while deleting 25\% of the internal nodes 
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having 
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary 
nodes per tree).
%                     name    leaves  unary       binary      poly
% 0   poly_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 1   poly_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 2   poly_fully_sfiied_0.05  200.0   0.000000    185.691468  6.393174
% 3   poly_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 4   poly_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 5   poly_fully_sified_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the 
effects on the likelihood across the three examples are not entirely
comparable. It is remarkable, however, how little effect introducing 
these polytomies has on the likelihood surface in this example. For 
the ARGs containing unary nodes, removing 25\% of the internal 
nodes has negligible effect, and even on the fully simplified 
ARG has a very minor effect on the minimum.

\subsection*{Piecewise decomposition over time}
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under the Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices:
    from 0 to 1,000 generations ago $N_e$ is 10,000;
    from 1,000 to 5,000 generations ago it is 5,000;
    and past that it is 10,000.
    Remaining parameters: genome length is 10kb, 20 diploid samples, and
    $r=2.5\times10^{-8}$. 
    There are 148 edges in the resulting ARG.
    A single $N_e$ value was estimated for each
    of the three slices of the ARG separated by
    the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. The number of edges per time slice
    (from most recent to oldest) was: 68, 57, and 23. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

Our final example explores decomposition of the likelihood
into separate components for different segments of time.
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, in which each recombination
event has an exact time,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the left or
right hand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.

% This approach therefore strengthens
% the inherent ability of ARGs to provide us with temporal
% resolution and context to a dataset of genetic variation.
% It also opens up the opportunity for more efficient optimization
% strategies during inference given that each slice of the ARG can be
% processed independently.
